The need for accuracy and smoothness in numerical simulations

### 2024-09-04 slides/ostrava/cckm-ppam-2024.tex
Revised slides for PPAM 2024

\begin{block}{Key questions}
  Does it matter
  \begin{enumerate}
  \item if the drag coefficient
    \begin{equation*}
      \nu \rightarrow C_D(\nu)
    \end{equation*}
    is smooth or not?
  \item if the event equation
    \begin{equation*}
      y(t) = 0
    \end{equation*}
    is solved accurately or not?
      \end{enumerate}
 \end{block}
}

@@

\frame{\frametitle{The secondary point of this talk}

  It might be possible to assert experimentally if
  \vspace{0.5cm}
  \begin{enumerate} \setlength\itemsep{1em}
  \item your functions are sufficiently smooth
  \item your equations are solved with sufficient accuracy
  \end{enumerate}
}

@@

  We validate our models by demonstrating that
  \begin{equation}
    P - T \approx 0
  \end{equation}
  is a good approximation, but
  \begin{equation}
    P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
  \end{equation}
  so we need to assert that
  \begin{align}
    |T- A_h| & \ll |P - \hat{A}_h| \\
    |A_h - \hat{A}_h| & \ll |P - \hat{A}_h|
  \end{align}
  before we are certain that
  \begin{equation}
    P - T \approx P - \hat{A}_h
  \end{equation}
  is a good approximation.
}

@@

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
If 
\begin{equation}
  E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
\end{equation}
then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
Why is this significant when the computer returns $\hat{A}_h$ and $$\hat{A}_h \not = A_h$$
}

@@

  We used GROMACS to simulate the motion of hen egg white lysozyme submerged in water.
  \begin{enumerate} \setlength\itemsep{1em}
  \item $T$: total energy of the simulation at the end
  \item $A_h$: the approximation of $T$ computed using SHAKE
  \item $\hat{A}_h$: the value of $A_h$ returned by the computer
  \end{enumerate}
\vspace{0.5cm}
  Temporal matters:
  \begin{enumerate}
  \item The length of the simulations was $10^{-12}$ s ($1$ ps)
  \item A common step size in MD is $10^{-15} s$ ($1$ fs)
  \end{enumerate}

@@

  \begin{figure}
    \includegraphics[width=12cm]{oplsaaltol12.pdf}
  \end{figure}
   The rapid growth of the energy for large value of $n$ 
    \begin{center}
      can likely be cured using compensated summation.
    \end{center}
  }

@@

  \begin{itemize}
  \item The \underline{wiggles} near $n=1000$ steps are a great concern.
  \item If there is an AEX, then $1$ fs is \underline{not} inside the asymp. range.
  \item We cannot quantifty the effect of changing the force-field.
  \end{itemize}

@@

\frame{\frametitle{What can kill an AEX?}

>>>>

\frame{\frametitle{Warning: Disturbing pictures to follow}

  We did two sets of simulations of a molecule in water
  \vspace{1cm}
  \begin{enumerate} \setlength\itemsep{1em}
    \item fixed real time
    \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
    \item identical initial conditions
   \end{enumerate}
  \vspace{1cm}
  and tracked the energy as a function of the number of steps.

\frame{\frametitle{Warning: Disturbing pictures below}

 \begin{figure}
    \includegraphics[width=8cm]{oplsaaltol12.pdf}
  \end{figure}

  \begin{figure}
    \includegraphics[width=8cm]{charmm36tol12.pdf}
  \end{figure}

%% \frame{\frametitle{External ballistics}

%%   \includegraphics[width=12cm]{range_rk1_mwe2.pdf}

%%   The total force $\bm{F}$ acting on the shell 
%%   \begin{equation}
%%     \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
%%   \end{equation}
%%   is the combination of gravity and aerodynamic drag.
%% }

%% \frame{\frametitle{External ballistics}

%% \begin{block}{Key questions}
%%   Does it matter
%%   \begin{enumerate}
%%   \item if the drag coefficient
%%     \begin{equation*}
%%       \nu \rightarrow C_D(\nu)
%%     \end{equation*}
%%     is smooth or not?
%%   \item if the event equation
%%     \begin{equation*}
%%       y(t) = 0
%%     \end{equation*}
%%     is solved accurately or not?
%%       \end{enumerate}
%%  \end{block}
%% }

@@

%% \frame{\frametitle{The secondary point of this talk}

%%   It might be possible to assert experimentally if
%%   \vspace{0.5cm}
%%   \begin{enumerate} \setlength\itemsep{1em}
%%   \item your functions are sufficiently smooth
%%   \item your equations are solved with sufficient accuracy
%%   \end{enumerate}
%% }

@@

  We validate our models by demonstrating that
  \begin{equation}
    P - T \approx 0
  \end{equation}
  is a good approximation, but
  \begin{equation}
    P - T \approx P - \hat{A}_h
  \end{equation}
  is not necessarily a good approximation. We have

  \begin{equation}
    P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
  \end{equation}
  so we need to assert that
  \begin{align}
    |T- A_h| & \ll |P - \hat{A}_h| \\
    |A_h - \hat{A}_h| & \ll |P - \hat{A}_h|
  \end{align}
}

@@

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
If 
\begin{equation}
  E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
\end{equation}
then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
which implies
\begin{equation}
  \log | 2^p - F_h| \approx \log|\text{constant}| + (q-p) \log h
\end{equation}

@@

  GROMACS simulation of hen egg white lysozyme in water:
  \vspace{0.25cm}
  \begin{enumerate} \setlength\itemsep{1em}
  \item $T$: total energy of the simulation at the end
  \item $A_h$: the approximation of $T$ computed using SHAKE
  \item $\hat{A}_h$: the value of $A_h$ returned by the computer
  \end{enumerate}
\vspace{0.5cm}
Temporal matters:
\vspace{0.25cm}
  \begin{enumerate}
  \item The length of the simulations was $10^{-12}$ s ($1$ ps)
  \item A common step size in MD is $10^{-15} s$ ($1$ fs)
  \end{enumerate}

@@

  \begin{figure}
    \includegraphics[width=10cm]{oplsaaltol12.pdf}
  \end{figure}
   The rapid growth of the energy for large value of $n$ 
    \begin{center}
      can likely be cured using compensated summation.
    \end{center}
  }

@@

  \begin{itemize}
  \item The \underline{wiggles} near $n=1000$ steps are a great concern.
  \item If there is an AEX, then $1$ fs is \underline{not} inside the asymp. range.
  \item We cannot assert that rounding errors are irrelevant
  \item We cannot estimate the discretication error
  \end{itemize}

@@

\frame{\frametitle{What can go wrong?}

  \begin{enumerate}\setlength\itemsep{1.5em}
  \item Proving the existence of an AEX is an exercise in
    \begin{center}
      Taylor expansion
    \end{center}
    We cannot complete the proof unless
    \begin{center}
      the functions are sufficiently smooth.
      \end{center}
    \item We cannot expect $\hat{A}_h$ to behave as predicted by the AEX unless
    $$ \hat{A}_h \approx A_h $$
    is a good approximation
    \end{enumerate}
}

\frame{\frametitle{Computing the optimal range of the a howitzer}

 \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk1_tol18.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and sufficiently accurate event location.}
  \end{figure}

\frame{\frametitle{Computing the optimal range of the a howitzer}

 \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk1_tol11.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and inaccurate event location.}
  \end{figure}

\frame{\frametitle{Modelling ions}
